Section Rings of $\bq$-divisors on minimal rational surfaces

### 2017-10-30 surfaces-write-up/surface-paper.tex
surfaces: address 29 & 31

Hence, for the remainder of the proof, we assume $D' = 0$.
%We will first check that $F^{\vec v}_i$ determine a system of minimal
%generators of $R_D$. We know $\left\{F_{1}^{\vec{v}} \colon \deg \vec{v} =
%\left\lfloor
%\frac{p}{q} \right\rfloor \right\}$ are minimal generators in
%degree $1$. Now let $i > 1$. Suppose $F_i^{\vec{v}}$ were not
%minimal for some $\vec{v} \in \mss_i$. Then $F_i^{\vec{v}} =
%F_{s}^{\vec{y}} F_{t}^{\vec{z}}$ for some $s, t \in \bn$ and
%$\vec{y}, \vec{z} \in \bz_{ \geq 0}^{m + 1}$. In particular, we see
%that 
%\begin{align*}
%	&d_i = s + t \\
%	&c_i = \deg \vec{v} = \deg \vec{y} + \deg \vec{z}
%\end{align*}
%
%\noindent
%so $\frac{c_i}{d_i}$ is the mediant of the rational numbers
%$\frac{\deg \vec{y}}{s}, \frac{\deg \vec{z}}{t}$ which both
%do not exceed $\alpha$. One of $\frac{\deg \vec{y}}{s},
%\frac{\deg \vec{z}}{t}$ must be at least $\frac{c_i}{d_i}$
%and $s$ and $t$ are both less than $d_i$, which contradicts the assumption that
%$\frac{c_i}{d_i}$ is a convergent of the Hirzebruch-Jung continued
%fraction of $\alpha$. Therefore, none of the $F_i^{\vec v}$ are redundant.
%
Next we show that $F^{\vec v}_i$ generate all of $R_D$. In the case $m = 1$, O'Dorney \cite[Theorem 6]
{dorney:canonical} demonstrates that each lattice point $(\beta, \gamma) \in
\bz_{\geq 0}^2$ with $\gamma \leq \beta \alpha$ lies in the $\bz_{\geq 0}$ span
of $(d_h, c_h)$ and $(d_{h + 1}, c_{h + 1})$ for
some $h \in \{0, \ldots, r\}$. A similar strategy works in the case $m > 1$. Let $(\beta, \gamma) = \kappa_1
(d_h, c_h) + \kappa_2 (d_{h + 1}, c_{h + 1})$ for $\kappa_1, \kappa_2 \in
\bz_{\geq 0}$. Any element $\frac{u^{\beta}
x^{\vec{v}}} {x_k^{ \gamma}} \in R_D$ is expressible as
\begin{align*}
	\frac{u^{\beta} x^{\vec{v}}} {x_k^{\gamma}} = \left(\frac{u^{d_h}}
	{x_k^{c_h}}\right)^{\kappa_1} \left(\frac{u^{d_{h + 1}}}
	{x_k^{c_{h + 1}}}\right)^{\kappa_2} x^{\vec{v}}.
\end{align*}

\noindent
We can then write $\vec{v}  = \sum_{\lambda=1}^{\kappa_1}\vec{w}_{(\lambda)} +
\sum_{\eta=1}^{\kappa_2} \vec z_{(\eta)}$ with $w_{(\lambda)} \in \mss_h$ and
$z_{(\eta)} \in \mss_{h+1}$ to give a decomposition
\begin{align}
\label{eqn:one-point-canonical-form}
	\frac{u^{\beta} x^{\vec{v}}} {x_k^{\gamma}}	= \prod_{\lambda = 1}
	^{\kappa_1} \frac{u^{d_h} x^{\vec{w}_{(\lambda)}}} {x_k^{c_h}}
	\prod_{\eta = 1}^{\kappa_2} \frac{u^{d_{h + 1}} x^{\vec{z}_{(\eta)}}}
	{x_k^{c_{h + 1}}}
\end{align}
\noindent
consisting of products of generators $F_h^{\vec{y}_{(\lambda)}}$
and $F_{h + 1}^{\vec{z}_{(\eta)}}$ which are in the form
prescribed in the theorem statement. Since we wrote an arbitrary monomial
$\frac{u^{\beta} x^{\vec{v}}}{x_k^\gamma} \in R_D$ as a product of generators,
this shows that $F_i^{\vec v}$ generate $R_D$.

@@

Now, we may apply the relations $L_{i, j}^{\vec{v}, \vec{w}}$
to any monomial in the $F_i^{\vec{v}}$'s involving
indices that differ by at most $1$ to produce the canonical form
\begin{align*}
	(F_i^{\vec{v}_{(1)}})^{g_{\vec{v}_{(1)}}} \cdots
	(F_i^{\vec{v}_{(\kappa_1)}})^{g_{\vec{v}_{(\kappa_1)}}}
	(F_{i + 1}^{\vec{w}_{(1)}})^{g_{\vec{w}_{(1)}}} \cdots
	(F_{i + 1}^{\vec{w}_{(\kappa_2)}})^{g_{\vec{w}_{(\kappa_2)}}}
\end{align*}
%% TODO: replace \kappa_1 and \kappa_2 with two letters or use prod notation

\noindent
where $\vec{v}_{(1)} \prec \vec{v}_{(2)} \prec \ldots \prec
\vec{v}_{(\kappa_1)} \prec \vec{w}_{(1)} \prec \ldots \prec
\vec{w}_{(\kappa_2)}$. Consequently, the relations of form
$G_{i, j}^{\vec{v}, \vec{w}}$ and of form $L_{i, j}^{\vec{v},
\vec{w}}$ generate all the relations among the $F_{i}^{\vec{v},
\vec{w}}$.
\end{proof}

@@

\begin{rem}
\label{rem:proj-two-points}
We cannot extend this result to the case when
$D$ is supported at two hypersurfaces with arbitrary rational (non-effective)
coefficients in the same manner that O'Dorney does for the
$\bp^1$ case \cite[Section 4]{dorney:canonical}. As will be shown
in Example ~\ref{eg:hyperplane}, the degrees of generation of the
section ring of a general $\bq$-divisor supported on two
hyperplanes cannot be bounded so tightly. The two-point
$\bp^1$ result leverages the fact that $\bp^1$ has precisely two
independent coordinates, so that two distinct integral subschemes
cannot represent equations of only $m$ of the $m+1$ coordinates.
\end{rem}

@@

>>>>

Hence, for the remainder of the proof, we assume $D' = 0$.
%We will first check that $F^{\vec v}_i$ determine a system of minimal
%generators of $R_D$. We know $\left\{F_{1}^{\vec{v}} \colon \deg \vec{v} =
%\left\lfloor
%\frac{p}{q} \right\rfloor \right\}$ are minimal generators in
%degree $1$. Now let $i > 1$. Suppose $F_i^{\vec{v}}$ were not
%minimal for some $\vec{v} \in \mss_i$. Then $F_i^{\vec{v}} =
%F_{s}^{\vec{y}} F_{t}^{\vec{z}}$ for some $s, t \in \bn$ and
%$\vec{y}, \vec{z} \in \bz_{ \geq 0}^{m + 1}$. In particular, we see
%that 
%\begin{align*}
%	&d_i = s + t \\
%	&c_i = \deg \vec{v} = \deg \vec{y} + \deg \vec{z}
%\end{align*}
%
%\noindent
%so $\frac{c_i}{d_i}$ is the mediant of the rational numbers
%$\frac{\deg \vec{y}}{s}, \frac{\deg \vec{z}}{t}$ which both
%do not exceed $\alpha$. One of $\frac{\deg \vec{y}}{s},
%\frac{\deg \vec{z}}{t}$ must be at least $\frac{c_i}{d_i}$
%and $s$ and $t$ are both less than $d_i$, which contradicts the assumption that
%$\frac{c_i}{d_i}$ is a convergent of the Hirzebruch-Jung continued
%fraction of $\alpha$. Therefore, none of the $F_i^{\vec v}$ are redundant.
%

Next we show that $F^{\vec v}_i$ generate all of $R_D$. In the case $m = 1$, O'Dorney \cite[Theorem 6]
{dorney:canonical} demonstrates that each lattice point $(\beta, \gamma) \in
\bz_{\geq 0}^2$ with $\gamma \leq \beta \alpha$ lies in the $\bz_{\geq 0}$ span
of $(d_h, c_h)$ and $(d_{h + 1}, c_{h + 1})$ for
some $h \in \{0, \ldots, r\}$. A similar strategy works in the case $m > 1$. Let $(\beta, \gamma) = \lambda
(d_h, c_h) + \kappa (d_{h + 1}, c_{h + 1})$ for $\lambda, \kappa \in
\bz_{\geq 0}$. Any element $\frac{u^{\beta}
x^{\vec{v}}} {x_k^{ \gamma}} \in R_D$ is expressible as
\begin{align*}
	\frac{u^{\beta} x^{\vec{v}}} {x_k^{\gamma}} = \left(\frac{u^{d_h}}
	{x_k^{c_h}}\right)^{\lambda} \left(\frac{u^{d_{h + 1}}}
	{x_k^{c_{h + 1}}}\right)^{\kappa} x^{\vec{v}}.
\end{align*}

\noindent
We can then write $\vec{v}  = \sum_{\tau=1}^{\lambda}\vec{w}_{(\tau)} +
\sum_{\eta=1}^{\kappa} \vec z_{(\eta)}$ with $\vec w_{(\tau)} \in \mss_h$ and
$\vec z_{(\eta)} \in \mss_{h+1}$ to give a decomposition
\begin{align}
\label{eqn:one-point-canonical-form}
	\frac{u^{\beta} x^{\vec{v}}} {x_k^{\gamma}}	= \prod_{\tau = 1}
	^{\lambda} \frac{u^{d_h} x^{\vec{w}_{(\tau)}}} {x_k^{c_h}}
	\prod_{\eta = 1}^{\kappa} \frac{u^{d_{h + 1}} x^{\vec{z}_{(\eta)}}}
	{x_k^{c_{h + 1}}}
\end{align}
\noindent
consisting of products of generators $F_h^{\vec{y}_{(\lambda)}}$
and $F_{h + 1}^{\vec{z}_{(\eta)}}$ which are in the form
prescribed in the theorem statement. Since we wrote an arbitrary monomial
$\frac{u^{\beta} x^{\vec{v}}}{x_k^\gamma} \in R_D$ as a product of generators,
this shows that the $F_i^{\vec v}$ generate $R_D$.

@@

Now, we may apply the relations $L_{i, j}^{\vec{v}, \vec{w}}$
to any monomial in the $F_i^{\vec{v}}$'s involving
indices that differ by at most $1$ to produce the canonical form
\begin{align*}
	\prod_{\tau=1}^\lambda (F_i^{\vec{v}_{(\tau)}})^{g_{\vec{v}_{(\tau)}}}
	\prod_{\eta=1}^\kappa (F_{i+1}^{\vec{v}_{(\eta)}})^{g_{\vec{v}_{(\eta)}}}
	%=	(F_i^{\vec{v}_{(1)}})^{g_{\vec{v}_{(1)}}} \cdots
	%(F_i^{\vec{v}_{(\lambda)}})^{g_{\vec{v}_{(\lambda)}}}
	%(F_{i + 1}^{\vec{w}_{(1)}})^{g_{\vec{w}_{(1)}}} \cdots
	%(F_{i + 1}^{\vec{w}_{(\kappa)}})^{g_{\vec{w}_{(\kappa)}}}
\end{align*}

\noindent
where $\vec{v}_{(1)} \prec \vec{v}_{(2)} \prec \ldots \prec
\vec{v}_{(\lambda)} \prec \vec{w}_{(1)} \prec \ldots \prec
\vec{w}_{(\kappa)}$. Consequently, the relations of form
$G_{i, j}^{\vec{v}, \vec{w}}$ and the relations of form
$L_{i, j}^{\vec{v}, \vec{w}}$ generate all the relations among the
$F_{i}^{\vec{v}, \vec{w}}$.
\end{proof}

@@

@@

\begin{rem}
\label{rem:proj-two-points}
We cannot extend Theorem~\ref{thm:proj-one-point} to the case when
$D$ is supported at two hypersurfaces with arbitrary rational (non-effective)
coefficients in the same manner that O'Dorney does for the
$\bp^1$ case \cite[Section 4]{dorney:canonical}. As shown
in Example ~\ref{eg:hyperplane}, the degrees of generation of the
section ring of a general $\bq$-divisor supported on two
hyperplanes cannot be bounded so tightly. The two-point
$\bp^1$ result leverages the fact that $\bp^1$ has precisely two
independent coordinates, so that two distinct integral subschemes
cannot represent equations of only $m$ of the $m+1$ coordinates.
\end{rem}
